---
title: "First steps with MongoDB"
output: html_notebook
---

As always happens, when we want to interact with a database, a driver is needed. Someone who translates our needs to the serving system. In order to interact with MongoDB from R, [mongolite](https://jeroen.github.io/mongolite/) package will help us doing so.

```{r}
library(mongolite)
```

Easy-peasy! Now let's connect to the database.

```{r}
m <- mongo()
```

As simple as that! It is worth noticing that by default it will connect to:

* Host: localhost
* Port: 27017
* Database: test
* Collection: test

So it should be empty so far. We can actually ask for databases when connected to the admin collection.

```{r}
admin <- mongo(db = "admin")
admin$run('{"listDatabases":1}')
```

Funnny... test database doesn't even exist! But we can ask for its collections.

```{r}
m$run('{"listCollections":1}')
```
We can even insert some information.

```{r}
m$insert('{"name" : "test"}')
```
Inserted! But where?

```{r}
m$run('{"listCollections":1}')
```
Wait the collection got created! And the database?

```{r}
admin$run('{"listDatabases":1}')
```

The database too! So the information is there.

```{r}
m$find()
```

So, lesson number 1: 

* Typos can get us into trouble when using MongoDB!

